---
title: "fflmediation: Identifying Feed Forward Loops Using Mediation Analysis"
author: Tessa Han
date: 06-27-2020
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ffl_mediation_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


In *Mediation Analysis to Identify Feed Forward Loop from Transcriptome Data* (2020), we introduce a new methodology that applies mediation analysis to predict feed forward loops (FFLs) in one biological group (the one-group method) or to predict FFLs unique to one of two biological groups (the two-group method). The *fflmediation* package contains functions for the one-group and two-group methods.

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ""
)
```

```{r setup}
library(fflmediation)
```

## One-group method

The one-group method predicts FFLs in one biological group. 

##### 1. Format data

The one-group method requires paired miRNA and mRNA expression data for *n* samples in a single biological group in the following format:

```{r, include = TRUE}
#sample miRNA expression data: miRNAs x samples (in this example: 3 x 50)
sample_mirna_expr

#sample mRNA expression data: mRNAs x samples (in this example 5 x 50)
sample_mrna_expr
```


##### 2. Obtain candidate FFLs

Candidate FFLs are assembled using databases of gene interactions (TargetScan v7.2, miRTarBase v7.0, miRDB v6.0, miRanda v3.0, TransmiR v2.0, TRRUST v2.0, and ENCODE). Integrating these databases generates a list of 25,240,974 candidate miRNA-FFLs (```candidate_ffls_mirna_all``` dataframe, created by binding ```candidate_ffls_mirna_all_part1``` and ```candidate_ffls_mirna_all_part2```; 25,240,974 x 9) and a list of 178,872 candidate TF-FFLs (```candidate_ffls_tf_all``` dataframe; 178872 x 10).

To obtain candidate miRNA-FFLs for analyses, subset the ```candidate_ffls_mirna_all``` dataframe, selecting for candidate miRNA-FFLs (rows) of interest. To obtain candidate TF-FFLs for analyses, subset the ```candidate_ffls_tf_all``` dataframe, selecting for candidate TF-FFLs (rows) of interest. 

```{r, include = TRUE}
#candidate_ffls_mirna_all
candidate_ffls_mirna_all <- rbind(candidate_ffls_mirna_all_part1, candidate_ffls_mirna_all_part2)
head(candidate_ffls_mirna_all)

#candidate_ffls_tf_all
head(candidate_ffls_tf_all)

#sample candidate miRNA-FFLs for analyses (a subset of the rows in candidate_ffls_mirna_all): 3 x 9
sample_candidate_ffls_mirna
```


##### 3. Apply one-group method

To apply the two-group method, use the ```predict_ffls_one_group``` function. Refer to the documentation of the ```predict_ffls_one_group``` function for the description and options of each argument. 

The following is an example for miRNA-FFLs. For TF-FFLs, change the ```candidate_ffls``` argument from a dataframe of candidate miRNA-FFLs to a dataframe of candidate TF-FFLs and the ```ffl_type``` argument from "miRNA" to "TF". 


```{r, include = TRUE}
#one-group method, miRNA-FFL example
sample_output_onegroup_mirna_ffls <- 
  predict_ffls_one_group(mirna_expr = sample_mirna_expr, 
                         mrna_expr = sample_mrna_expr,
                         ffl_type = "miRNA",
                         candidate_ffls = sample_candidate_ffls_mirna,
                         num_bootstrap_samples = 100, num_permutations = 100,
                         p_value_adjust_method = "fdr", 
                         seed = 12345, 
                         parallel = FALSE)

sample_output_onegroup_mirna_ffls
```


## Two-group method

The two-group method predicts FFLs unique to one of two biological groups. 

##### 1. Format data

The two-group method requires paired miRNA and mRNA expression data for *n* samples in one biological group and *m* samples in another biological group in the following format:

```{r, include = TRUE}
#####group 1
#sample miRNA expression data: miRNAs x samples (in this example: 3 x 50)
sample_mirna_expr_g1

#sample mRNA expression data: mRNAs x samples (in this example 5 x 50)
sample_mrna_expr_g1

#####group 2
#sample miRNA expression data: miRNAs x samples (in this example: 3 x 50)
sample_mirna_expr_g2

#sample mRNA expression data: mRNAs x samples (in this example 5 x 50)
sample_mrna_expr_g2
```


##### 2. Obtain candidate FFLs

Obtain candidate FFLs for the two-group method in the same way as for the one-group method detailed above. 

##### 3. Apply two-group method

To apply the two-group method, use the ```predict_ffls_two_group``` function. Refer to the documentation of the ```predict_ffls_two_group``` function for the description and options of each argument. 

The following is an example for miRNA-FFLs. For TF-FFLs, change the ```candidate_ffls``` argument from a dataframe of candidate miRNA-FFLs to a dataframe of candidate TF-FFLs and the ```ffl_type``` argument from "miRNA" to "TF". 


```{r, include = TRUE}
#two-group method, miRNA-FFL example
sample_output_twogroups_mirna_ffls <- 
  predict_ffls_two_groups(mirna_expr_g1 = sample_mirna_expr_g1, 
                          mrna_expr_g1 = sample_mrna_expr_g1,
                          mirna_expr_g2 = sample_mirna_expr_g2, 
                          mrna_expr_g2 = sample_mrna_expr_g2,
                          ffl_type = "miRNA",
                          candidate_ffls = sample_candidate_ffls_mirna,
                          num_bootstrap_samples = 100, 
                          num_permutations = 100,
                          p_value_adjust_method = "fdr", 
                          seed = 12345,
                          parallel = FALSE)
sample_output_twogroups_mirna_ffls
```



